\LARGE{\bf Parallelizing a Network Simulator}

### 2009-03-16 doc/paper.tex
Added a figure for the lfqueue implementation.
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1157 ce7aba0d-9f81-40a6-a495-cecae84b0515

\subsubsection{Implementation}
Lock-free queues are based on the universal lock-free algorithm
\cite{LF} that assumes that any access to the shared data structure can
see the data structure in any potential partially updated state.  The
algorithm then allows for the access to help other accesses to
complete before actually completing itself.  While this may sound
conceptually simple, the combinatorial number of states of the data
structure and different readers and writers is astounding.\\
~\\
The lock-free queue implementation that we use begins by having a sentinel
at the head of the queue whose value is meaningless.  This eases some of
the implementation details of the queue.  Initially both the head and tail
pointers of the queue point to the sentinel.  Whenever a new item is added
to the queue, the tail pointer is moved to point to the end of the list.
Whenever an item is popped from the queue, the head is moved to point to
the new sentinel, the next item in the list.  In this case the old
sentinel can then be reclaimed.  Each element in the lock-free queue is
wrapped in an AtomicReference object.  This object ensures that the any
items in the queue can only be written to atomically.  A diagram of how
this works can be seen in Figure (need figure here).\\
~\\
We initially implemented this AtomicReference object with a PThread
lock.  While this provided a finer granularity of access to shared
objects than previously, we realized that the overhead in the number of
locks was tremendous.  Instead we relied upon an extension to GCC 4.1.2
that provided an atomic \_\_compare\_and\_exchange operation.
This operation was significantly less costly in terms of synchronization
overhead and was also much faster.\\
~\\
After testing our lock-free queues, we replaced all of
the queues in the simulator with our lock-free queue implementation.  While this
proved fairly straight forward, we didn't initially account for the high
overhead that the lock-free queues themselves created.  The lock-free
queues involve the creation of multiple objects and accessing the
lock-free queues involves chasing many pointers through these structures
as well as rearranging many pointers.  In some cases this overhead proved
to be extreme.  In order to remove it, we created a common interface that
was shared between the lock-free queues and the STL queue.  This then
allowed us to create dynamic instances of either a shared queue or an STL
queue depending on whether the channel was shared or not.  This significantly
improved our performance as we will demonstrate in section \ref{results}.

>>>>

\subsubsection{Implementation}
Lock-free queues are based on the universal lock-free algorithm
\cite{LF} that assumes that any access to the shared data structure can
see the data structure in any potential partially updated state.  The
algorithm then allows for the access to help other accesses to
complete before actually completing itself.  While this may sound
conceptually simple, the combinatorial number of states of the data
structure and different readers and writers is astounding.\\
~\\
The lock-free queue implementation that we use begins by having a sentinel
at the head of the queue whose value is meaningless.  This eases some of
the implementation details of the queue.  Initially both the head and tail
pointers of the queue point to the sentinel.  Whenever a new item is added
to the queue, the tail pointer is moved to point to the end of the list.
Whenever an item is popped from the queue, the head is moved to point to
the new sentinel, the next item in the list.  In this case the old
sentinel can then be reclaimed.  Each element in the lock-free queue is
wrapped in an AtomicReference object.  This object ensures that the any
items in the queue can only be written to atomically.  A diagram of how
this works can be seen in Figure \ref{lfqueue}.
\begin{figure}[h]
\centering
\includegraphics[width=3in, height=2.25in]{lfqueue.eps}
\caption{This figure demonstrates the layout of the objects associated with a lock-free queue of length 2 (the sentinel value does not count to towards the length of the list).  Note that the pointer to the next element in the list, does not point directly to next list node, but instead to the wrapper AtomicReference class.  Also note that the sentinel is not an actual maintained pointer, but will always point to the same AtomicReference as the Head Node.  It is included in this diagram for clarity.}
\label{lfqueue}
\end{figure}
We initially implemented this AtomicReference object with a PThread
lock.  While this provided a finer granularity of access to shared
objects than previously, we realized that the overhead in the number of
locks was tremendous.  Instead we relied upon an extension to GCC 4.1.2
that provided an atomic \_\_compare\_and\_exchange operation.
This operation was significantly less costly in terms of synchronization
overhead and was also much faster.\\
~\\
After testing our lock-free queues, we replaced all of
the queues in the simulator with our lock-free queue implementation.  While this
proved fairly straight forward, we didn't initially account for the high
overhead that the lock-free queues themselves created.  The lock-free
queues involve the creation of multiple objects and accessing the
lock-free queues involves chasing many pointers through these structures
as well as rearranging many pointers.  In some cases this overhead proved
to be extreme.  In order to remove it, we created a common interface that
was shared between the lock-free queues and the STL queue.  This then
allowed us to create dynamic instances of either a shared queue or an STL
queue depending on whether the channel was shared or not.  This significantly
improved our performance as we will demonstrate in section \ref{results}.
